Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-10 article/article.tex
locations
peace

  	\caption{Above is the hypothetical neural response to some stimulus. Below is the BOLD response that is evoked by the short spike of neural activity. The neural activity shown above is attempted to be inferred from the BOLD response that is measured by the fMRI scanner. Figure is adopted from Heeger and Ress \cite{he:2002na}.}

@@

\indent The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g. a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researches where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.\\

@@

\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROI's are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.

@@

To reducde the dimensionality of the input, two different methods are used.

ROI mapping. ROI mapping is a form of feature selection, wherein only a subset of the d-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. To reduce the number of features, average voxel values within this regions are taken as input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. ROIs require \textit{a priori} knowledge. When this knowledge is inaccurate, it results in a inductive bias.

>>>>

  	\caption{Above is the hypothetical neural response to some stimulus. Below is the BOLD response that is evoked by the short spike of neural activity. The neural activity shown above is attempted to be inferred from the BOLD response below that is measured by the fMRI scanner. Figure is adopted from Heeger and Ress \cite{he:2002na}.}

@@

\indent The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g. a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researchers where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.\\

@@

\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROIs are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.

@@

To reducde the dimensionality of the input, several different methods were used.

ROI mapping. ROI mapping is a form of feature selection, wherein only a subset of the d-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant \cite{wa:2003betw}. To further reduce the number of features, average voxel values within this regions are taken as new input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, it results in an inductive bias.
